---
title: "Assessment Data Responses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{assessment-data-responses}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Types of Response

We have designed the package to return two 'types' of assessment
response from the API, which we refer to as 'minimal' and 'full'
assessment responses.

## Minimal Responses

A 'minimal' response returns a `tibble()` containing just enough
information for the user to gather full assessment data for each
assessment they are interested in. For example, you may be interested 
in knowing all assessments which have habitat code 1 coded against them:

```{r eval=FALSE}
a <- assessments_by_group(api, "habitats", "1")

# A tibble: 114 × 7
   sis_taxon_id assessment_id latest year_published scopes_description_en scopes_code url                                               
          <int>         <int> <lgl>  <chr>          <chr>                 <chr>       <chr>                                             
 1       133958       3881104 TRUE   2008           Global                1           https://www.iucnredlist.org/species/133958/3881104
 2       134059       3887925 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134059/3887925
 3       134019       3893117 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134019/3893117
 4       134151       3910739 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134151/3910739
 5       134199       3918266 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134199/3918266
 6       134378       3947233 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134378/3947233
 7       134409       3952097 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134409/3952097
 8       134409       3952097 TRUE   2008           Eastern Africa        18          https://www.iucnredlist.org/species/134409/3952097
 9       134409       3952097 TRUE   2008           Pan-Africa            10          https://www.iucnredlist.org/species/134409/3952097
10       134438       3956693 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134438/3956693
# ℹ 104 more rows
# ℹ Use `print(n = ...)` to see more rows
```

With no query parameters (arguments) passed to `assessments_by_group()`
you'll get the above dataframe with a row for each unique assessment ID
in the Red List database for all scopes and years. At this point you may choose 
to filter out just those assessments which have a scope of 'Eastern Africa' or 
'Pan-Africa':

```{r eval=FALSE}
ea <- a %>% 
        dplyr::filter(scopes_code == 18 | scopes_code == 10)
```

You could then use this 'minimal' dataframe and loop over each assessment ID to 
gather 'full' assessment data for each assessment:

```{r eval=FALSE}
all_data <- assessment_data_many(api, ea$assessment_id)
```

If you know _a priori_ which assessments you need, you can supply the appropriate 
arguments to `assessments_by_group()` directly, and save a filtering task further
downstream (it will also likely save on calls to the API database):

```{r eval=FALSE}
# Get global assessments published in 2008 for habitat code 1
assessments_by_group(api, "habitats", "1", year_published = 2008, scope = 1)

# A tibble: 38 × 7
   sis_taxon_id assessment_id latest year_published scopes_description_en scopes_code url                                               
          <int>         <int> <lgl>  <chr>          <chr>                 <chr>       <chr>                                             
 1       133958       3881104 TRUE   2008           Global                1           https://www.iucnredlist.org/species/133958/3881104
 2       134059       3887925 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134059/3887925
 3       134019       3893117 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134019/3893117
 4       134151       3910739 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134151/3910739
 5       134199       3918266 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134199/3918266
 6       134378       3947233 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134378/3947233
 7       134409       3952097 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134409/3952097
 8       134409       3952097 TRUE   2008           Eastern Africa        18          https://www.iucnredlist.org/species/134409/3952097
 9       134409       3952097 TRUE   2008           Pan-Africa            10          https://www.iucnredlist.org/species/134409/3952097
10       134438       3956693 TRUE   2008           Global                1           https://www.iucnredlist.org/species/134438/3956693
# ℹ 28 more rows
# ℹ Use `print(n = ...)` to see more rows
```

### Minimal assessment data functions

Minimal assessment data is returned from the assessments_by_ functions:

- `assessments_by_group()`
- `assessments_by_name()`
- `assessments_by_sis_id()`
- `assessments_by_taxonomy()`

## Full Responses
In contrast to the 'minimal' responses, you can get 'full' assessment data by first 
calling `assessment_data()` to get all assessment data in 'raw' format:

```{r eval=FALSE}
# Get full assessment data for the black rhino
a <- assessment_data(api, 152728945)
```

The above function returns the full response from the API, with no transformations
(except coercing JSON to lists). You can then call `parse_assessment_data()` to
get a tidied named list of tibbles:

```{r eval=FALSE}
a <- parse_assessment_data(a)
```

When calling `parse_assessment_data()` you are allowing the {iucnredlist} package
to perform a number of data tidying routines. In a nutshell this function will:

1. Tidy the 'taxon' element in the raw response, moving Common Names, SSC Groups 
and Synonyms to named elements at the top-most level in the list, prepending 
these parts of the response with `taxon_`.
2. Move conservation_actions_in_place from under conservation_actions 
to the top-most level
3. Remove the 'documentation' part of the response. as this is mostly long paragraphs 
of text and HTML. If part of the response is important for your work, we would 
recommend fetching this from the object created with `assessment_data()` and 
process according to your needs.
4. Ordering the named lists alphabetically

### Full assessment data functions
Full assessment data is returned from the functions:

- `assessment_data()`
- `assessment_data_many()`


