---
title: "Gather Common Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gather-common-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
This tutorial has been created to walk you through gathering common data from the
IUCN Red List API. In Part 1, we will stay completely within the API by getting a 
list of species' assessments directly from the taxon endpoint. In Part 2, we will
show you how to import your own list of species' names and perform the same data
gathering steps.

## Part 1 - Gather habitats & threats for the _felidae_
First, you'll want to load the package into your R session and then initialise
a connection to the Red List API. Initialising a connection will require you to have
a valid API token. If you don't already have one, you can sign up for one 
[here](https://api.iucnredlist.org). 

```{r eval = FALSE}
library(iucnredlist)
api <- init_api("your_api_token")
```





